---
title: "Clustering experimental macrophage gene expression data"
author: "mleukam"
date: "2019-06-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Clear workspace prior to analysis and load necessary packages
```{r}
rm(list = ls())
```

```{r}
library("tidyverse")
library("factoextra")
library("WGCNA")
library("annotate")
library("lumi")
library("FactoMineR")
library("ggsci")
library("ggpubr")
library("DESeq2")
```

Load data
```{r}
mac_eset <- readRDS("~/tcga_macs/output/mac_eset.rds")
lumi_46903 <- readRDS("~/tcga_macs/output/normalized_probe_intensities.Rds")
```

#### Filter probes

Detailed instructions for building a contrast matrix and filtering low-expression scores can be found in the [Limma User's Guide](http://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf), particularly chapters 9.2 and 9.3

```{r}
# retreive normalized data
dataMatrix <- exprs(mac_eset)

# remove the probes that are not detected in raw data
presentCount <- detectionCall(lumi_46903)
selDataMatrix <- dataMatrix[presentCount > 0,]
probeList <- rownames(selDataMatrix)
```

## PCA

To get a sense of the relationships among the samples, we will start with a simple PCA plot labeled by treatment and then by the clusters derived in the original paper.

Source: https://rpkgs.datanovia.com/factoextra/index.html

```{r}
# transpose matrix to tidy data format
t_DataMatrix <- t(selDataMatrix)
t_DataMatrix[1:5, 1:5]

# extract phenotype data
pheno_data <- pData(mac_eset)

# add categories and factors
# select columns of interest from phenotype data
pheno_pca <- pheno_data %>%
  rownames_to_column(var = "raw_id") %>%
  dplyr::select(raw_id, treatment, clusternum)

df_pca <- t_DataMatrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "raw_id") %>%
  left_join(pheno_pca, by = "raw_id") %>%
  as_tibble() %>%
  dplyr::select(raw_id, treatment, clusternum, everything()) %>%
  print()

norm_probe_pca <- PCA(df_pca[,c(4:39471)], graph = FALSE)
head(get_eig(norm_probe_pca))
fviz_eig(norm_probe_pca)
```

Source for overriding shape change with habillage argument: https://github.com/kassambara/factoextra/issues/20

```{r}
# visualize eigenvalues
mypal <- pal_igv("default", alpha = 1)(23)
pca1 <- fviz_pca_ind(norm_probe_pca,
             label = "none", # hide individual labels
             habillage = "none",
             col.ind = df_pca$treatment,
             palette = mypal,
             geom.ind = "point",
             invisible = "quali",
             pointshape = 19,
             title = "PCA of normalized expression"
) +
  labs(subtitle = "By activation agent")
print(pca1)

pca2 <- fviz_pca_ind(norm_probe_pca,
             label = "none", # hide individual labels
             habillage = "none",
             col.ind = df_pca$clusternum,
             palette = mypal,
             geom.ind = "point",
             invisible = "quali",
             pointshape = 19, 
             addEllipses = TRUE,
             title = "PCA of normalized expression"
) +
  labs(subtitle = "By cluster assignment")
print(pca2)
```


## Correlation Coefficient Matrix Analysis

Background and references:
```
Method source for correlation matrix: https://davetang.org/muse/2012/01/31/creating-a-correlation-matrix-with-r/

More on constructing a correlation matrix:
https://www.researchgate.net/post/How_to_calculate_Gene-Gene_Pearson_correlation

WGCNA package: https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/index.html
```

Following Xue et al Immunity (2014), select the 1000 probes with the highest variance to reduce the computational load to make a correlation matrix

Source for selecting highest variance: http://r.789695.n4.nabble.com/Help-on-selecting-genes-showing-highest-variance-td3580645.html

```{r}
# select 1000 highest probes with highest variation
# input is normalized expression matrix with zero expression genes removed
myvars <- apply(selDataMatrix, 1, var, na.rm = TRUE) 
myvars <- sort(myvars, decreasing = TRUE) 
myvars <- myvars[1:1000] 
data_matrix_onek <- selDataMatrix[names(myvars), ] 
dim(data_matrix_onek) 
```

#### Heirarchical Clustering

Quick hierarchical clustering to look for outliers. Will also visualize treatment condition and cluster assignment from the original paper.

```{r}
# heirarchical cluster
datExpr <- t(data_matrix_onek)
sampleTree = hclust(dist(datExpr), method = "average")
sizeGrWindow(16,9)
par(cex = 0.4)
plot(sampleTree, main = "Sample clustering to detect outliers", 
     sub = "", 
     xlab = "", 
     cex.lab = 1.5,
     cex.axis = 1.5, 
     cex.main = 2.5)

# no significant outliers
```

Set up phenotype data for WGCNA
```{r}
# use pheno_data from above with treatment identification and cluster number
# select columns I want to visualize
traitData <- pheno_data %>%
  dplyr::select(treatment, clusternum) %>%
  print()

treat_colors <- pal_igv("default", alpha = 1)(23)
names(treat_colors) <- levels(traitData$treatment)
treat_colors <- enframe(treat_colors) %>%
  dplyr::rename(treatment = name, treat_colors = value) %>%
  mutate(treatment = as.factor(treatment))
clust_colors <- pal_igv("default", alpha = 1)(9)
names(clust_colors) <- levels(traitData$clusternum)
clust_colors <- enframe(clust_colors) %>%
  dplyr::rename(clusternum = name, clust_colors = value) %>%
  mutate(clusternum = as.factor(clusternum))

traitColors <- pheno_data %>% 
  left_join(treat_colors, by = "treatment") %>%
  left_join(clust_colors, by = "clusternum") %>%
  dplyr::select(treat_colors, clust_colors) %>%
  print()

```

```{r}
sampleTree2 <- hclust(dist(datExpr), method = "average")
plotDendroAndColors(sampleTree2, traitColors,
                    groupLabels = names(traitData), 
                    main = "Sample dendrogram and treatment assignments")

```


#### Correlation matrix for individual samples

Will also use top 1000 most variable probes for correlation clustering to limit memory usage and eliminate noise.

Source for plots: https://www.datanovia.com/en/blog/cluster-analysis-in-r-simplified-and-enhanced/


```{r}
# create correlation matrix for each sample 
# using 1000 most variable probes as defined above
res_dist <- get_dist(datExpr, method = "pearson")
head(round(as.matrix(res_dist), 2))[, 1:6]

# plot dissimilarity matrix
fviz_dist(res_dist, lab_size = 8)
```

Hierarchical and K-means clustering of correlations
```{r}
# Save correlation matrix as variable
res_dist_m <- as.matrix(res_dist)

# Compute hierarchical clustering
res_hc <- hclust(res_dist, method = "ward.D2")

# Visualize
plot(res_hc, cex = 0.5)

# Enhanced k-means clustering
res_km <- eclust(res_dist_m, "kmeans", nstart = 25)
```

Diagnostic plots
```{r}
# Gap statistic plot
fviz_gap_stat(res_km$gap_stat)

# Silhouette plot
fviz_silhouette(res_km)

# Optimal number of clusters using gap statistics
res_km$nbclust

# Print result
res_km
```

#### Combining cluster assignments with phenodata

```{r}
# extract cluster assignment from km results
clust_asgn <- enframe(res_km$cluster) %>%
  dplyr::rename(rawID = name, km_clust = value)

# combine results with phenotype data
# compare cluster assignments to treatments and original clusters
pheno_tbl <- pData(mac_eset) %>%
  rownames_to_column(var = "rawID") %>%
  as_tibble() %>%
  left_join(clust_asgn, by = "rawID") %>%
  arrange(treatment) %>%
  dplyr::select(rawID, treatment, km_clust, clusternum) %>%
  print()

write_csv(pheno_tbl, "~/tcga_macs/output/km_cluster_assignments.csv")
```

#### Correlation of collapsed treatment groups

Data preparation
```{r}
# format data so probes are in rows and samples are in columns
# will use only top 1000 most variable genes
data_matrix_onek[1:5, 1:5]

# pull out a table of treatment and samples names from pheno_data
txgrp <- pheno_tbl %>%
  dplyr::select(rawID, treatment)

group_df <- colnames(data_matrix_onek) %>%
  enframe() %>%
  dplyr::rename(rawID = value) %>%
  left_join(txgrp)

# check that order of samples matches the data matrix perfectly
identical(colnames(data_matrix_onek), group_df$rawID)

# pull treatment assignment as a vector
group_vect <- group_df$treatment

# format data so probes are in columns and samples are in rows
tx_row_matrix <- t(data_matrix_onek)
tx_row_matrix[1:5, 1:5]

# check the order of samples still matches the treatment source perfectly
identical(rownames(tx_row_matrix), group_df$rawID)

# use WGCNA function for collapsing identical probes
tx_collapsed <- collapseRows(tx_row_matrix, 
                          rowGroup = group_vect, 
                          rowID = rownames(tx_row_matrix), 
                          method = "Average")

# pull out expression matrix by treatment
tx_matrix <- tx_collapsed$datETcollapsed
nrow(tx_matrix)
tx_matrix[1:5, 1:5]
```

Correlation matrix by treatment group
```{r}
# create correlation matrix for each sample 
# using 1000 most variable probes as defined above
res_dist_tx <- get_dist(tx_matrix, method = "pearson")
head(round(as.matrix(res_dist), 2))[, 1:6]

# plot dissimilarity matrix
fviz_dist(res_dist_tx, lab_size = 8)
```

Hierarchical and K-means clustering of treatment group correlations
```{r}
# Save correlation matrix as variable
res_dist_tx_m <- as.matrix(res_dist_tx)

# Compute hierarchical clustering
res_hc_tx <- hclust(res_dist_tx, method = "ward.D2")

# Visualize
plot(res_hc_tx, cex = 0.5)

# Enhanced k-means clustering
res_km_tx <- eclust(res_dist_tx_m, "kmeans", nstart = 25)
```

Diagnostic plots
```{r}
# Gap statistic plot
fviz_gap_stat(res_km_tx$gap_stat)

# Silhouette plot
fviz_silhouette(res_km_tx)

# Optimal number of clusters using gap statistics
res_km_tx$nbclust

# Print result
res_km_tx
```

Other values for k - visual review
```{r}
res_km_4  <- eclust(res_dist_tx_m, "kmeans", k = 4)
res_km_5  <- eclust(res_dist_tx_m, "kmeans", k = 5)
res_km_6  <- eclust(res_dist_tx_m, "kmeans", k = 6)
res_km_7  <- eclust(res_dist_tx_m, "kmeans", k = 7)
res_km_8  <- eclust(res_dist_tx_m, "kmeans", k = 8)
res_km_9  <- eclust(res_dist_tx_m, "kmeans", k = 9)
res_km_10 <- eclust(res_dist_tx_m, "kmeans", k = 10)
```

K = 7 best approximates the different known biological subtypes, and most closely resembles the clusters derived from the original paper.

## Summary and diagnostic plots

Silhouette plot of km = 7 model
```{r}
# Silhouette plot
fviz_silhouette(res_km_7)
```

PCA plot using all gene expression data for individual samples, now with new cluster labels applied
```{r}
# Get cluster assignments for each condition and add to phenodata
myclust <- enframe(res_km_7$cluster) %>%
  dplyr::rename(treatment = name, 
                mycluster = value)

pheno_tbl_clust <- pheno_tbl %>% 
  left_join(myclust, by = "treatment") %>%
  mutate(treatment = as.factor(treatment),
         mycluster = as.factor(mycluster)) %>%
  dplyr::rename(sample_cluster = km_clust,
         treatment_cluster = mycluster,
         orig_cluster = clusternum) %>%
  mutate(sample_cluster = as.factor(sample_cluster)) %>%
  mutate(adj_cluster = ifelse(treatment == "control", 0, treatment_cluster)) %>%
  mutate(adj_cluster = as.factor(adj_cluster)) %>%
  print()

# use normalized gene expression values from PCA plots above
# t_DataMatrix is gene expression matrix in tidy format from earlier in this notebook
df_pca_clust <- t_DataMatrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "rawID") %>%
  left_join(pheno_tbl_clust, by = "rawID") %>%
  as_tibble() %>%
  dplyr::select(rawID, treatment, adj_cluster, treatment_cluster, sample_cluster, orig_cluster, everything()) %>%
  print()

pca_clust <- PCA(df_pca_clust[,c(7:39474)], graph = FALSE)
head(get_eig(pca_clust))
fviz_eig(pca_clust)

mypal <- pal_igv("default", alpha = 1)(23)

pca3 <- fviz_pca_ind(pca_clust,
             label = "none", # hide individual labels
             habillage = "none",
             col.ind = df_pca_clust$adj_cluster,
             palette = mypal,
             geom.ind = "point",
             invisible = "quali",
             pointshape = 19, 
             addEllipses = TRUE,
             title = "PCA of normalized expression"
) +
  labs(subtitle = "By treatment-level cluster")
print(pca3)

```

## Remake expression set for export
```{r}

# add cluster information to phenodata
pheno_data_3 <- pheno_data %>%
  rownames_to_column(var = "rawID") %>%
  as_tibble() %>%
  left_join(pheno_tbl_clust)

# add back rownames to phenodata
mac_pData <- pheno_data_3 %>%
  as.data.frame() %>%
  column_to_rownames(var = "rawID")

# create a frame for annotation of column names in metadata
# will be empty for now
varnames <- colnames(mac_pData)
vardata <- data.frame(labelDescription = varnames, 
                      row.names = varnames)

# rebuild AnnotatedDataFrame
mac_phenoData_3 <- new("AnnotatedDataFrame",
                       data = mac_pData, 
                       varMetadata = vardata)

# retreive expression data matrix
mac_exprs_3 <- exprs(mac_eset)

# create expression set
mac_eset_3 <- ExpressionSet(assayData = mac_exprs_3,
                            phenoData = mac_phenoData_3,
                            annotation = "Illumina Human HT-12 v3.0 beadchip")

# check properties
featureNames(mac_eset_3)[1:5]
sampleNames(mac_eset_3)[1:5]
varLabels(mac_eset_3)
mat <- exprs(mac_eset_3)
dim(mat)
mat[1:5, 1:5]

# save expression set
saveRDS(mac_eset_3, "~/tcga_macs/output/mac_eset_w_clusters.rds")
```
